Monoidal Toposes of Grammars and Semantic Actions

### 2022-08-14 notes.tex
eureka!?

\section{Language Combinators in Monoidal Toposes}

@@

@@

\subsection{Internal Language of Monoidal Toposes}

A powerful tool that toposes provides is the use of \emph{internal
languages} which in the case of a topos can interpret all of
intuitionistic set theory or type theory. When working with specific
toposes it is fruitful to extend the type theory with new axioms and
structures based on the structure present in the category.

For a \emph{monoidal topos} this would be some form of
\emph{separation type theory}, a proof-relevant form of ordered
separation logic. Is there any syntax we can use off the shelf?

Since the topos of semantic actions is constructed by Artin gluing, it
contains the category of $\Set$ as an open subtopos and $\Grammar$ as
its closed complement. This structure can be axiomatized as an
abstract proposition $\Phi$ which we can think of as ``when $\Phi$ is
true we are in the subtopos $\Set$.''  Sterling and Harper have termed
this situation a ``phase distinction''. From $\Phi$, one can define
open and closed modalities $\fullmoon$ and $\newmoon$ that ``project
out'' the underlying set and grammar from a semantic action.
%
Then we might be able to internally describe the pushforward of an
action $(a,g,X)$ along a function $f : X \to Y$ as follows. First we
assume we have a set $A$ (corresponding to $(a,g,X)$) and a
$\Phi$-modal set $Y$ (corresponding to $Y$). Then given any $f : A \to
Y$ we can define the pushforward as the refinement type
\[ [ \Phi \hookrightarrow y : Y | \newmoon (\sum_{x:A} f(x) = b) ]\]

@@

>>>>

\section{Grammar Combinators in Monoidal Toposes}

@@

\section{Internal Language for the Topos of Semantic Actions}

A powerful tool for expressing constructions in a topos is the use of
the \emph{internal language}, which for a generic topos is a form of
extensional dependent type theory.
%
To extract maximum benefit from the internal language, we add axioms
and judgments to type theory for the specific topos models we are
working with.
%
For the toposes of grammars and semantic actions we have a plethora of
structure, we highlight the three most relevant:

\begin{itemize}
\item Both are toposes so models of extensional dependent type theory
\item They are models of the (non-dependent) ordered logic of bunched
  implications in that they are non-symmetric monoidal closed as well
  as Cartesian closed.
\item The topos of semantic actions, being constructed by Artin
  gluing, contains a proposition $\Phi$ that presents the categories
  $\Set$ and $\Grammar$ as open and complementary closed
  subtoposes. Sterling and Harper refer to this situation as a ``phase
  distinction''.
\end{itemize}

Combining the internal language of dependent type theory with bunched
implications or even just monoidal structure is tricky. The reason is
that all constructions need to be stable under pullback to be used in
dependent type theory, but our monoidal structure is not stable under
pullback, as is typical.
%
Fortunately, the phase distinction allows for a fruitful combination
of these aspects, called \emph{quantitative type theory}
\cite{atkey-qtt,mcbride-nuttin}.
%
We will need a non-symmetric variant.

@@

@@

\bibliographystyle{plain}
\bibliography{grammar}
